%!TEX root = ../../thesis.tex
\null
\vfill

\begin{center}

{\LARGE{\scshape Acknowledgements}}\\

\noindent\rule[2pt]{0.85\textwidth}{0.5pt}

\parbox{0.85\textwidth}{
This thesis is the culmination of over four years of collaborative team work. I
am very grateful to many people that helped me along the way.\\

First and foremost, my advisor and partner in crime Bernardo, who was with me
every step of the way for over four years. His guidance, support, and friendship
will always be appreciated. I hope we will have lots more fun in the years to
come.\\

I extend a tremendous thanks to my predoctoral and doctoral committees, Dr.
Francisco Villa Villa, Dr. Norberto Arzate Plata, Dr. Ram\'on Carriles Jaimes,
and Dr. Wolf Luis Moch\'an Backal. Their expediency, comments, and suggestions
were invaluable for this work.\\

I am very grateful to all my colleagues at the Laboratoire des Solides
Irradi\'es in France. In particular, Nicolas Tancogne-Dejean and Val\'erie
V\'eniard for their excellent guidance and hard work on this project. I have
them to thank for two publications, with more to come.\\

I want to thank all my friends at the CIO for their friendship and support. In
particular, Ana Luisa Aguayo Alvarado for her kindness and warmth, Mario ``el
ticher'' Ruiz Berganza for his jokes and guidance, Reinaldo Arturo Zapata Pe\~na
and his lovely family for welcoming me as one of their own, and Sergio Augusto
Romero Serv\'in for always being there. Thank you all for many years of
adventures and fun.\\

I am grateful towards the CONACyT and the CIO for the financial support and
infrastructure contributed towards this work.\\

I also wish to thank my parents, Mike and Ana, for the endless support and love
that they have given me throughout the years. They have always supported my
endeavors in all possible ways.\\

Lastly, I want to thank my wife Lili for her love, support, and patience during
these months apart. The promise of a wonderful future with my new family has
kept me focused even in the darkest of moments. This is the final stepping into
the first chapter of our lives together.}

\noindent\rule[2pt]{0.85\textwidth}{0.5pt}

\end{center}

\vfill

\cleardoublepage
